Forcing for Isabelle

### 2018-12-19 LICS/easy-axioms.tex
ed. varias, esp. Hacking y comentario para Miguel

The Union Axiom asserts that if $x$ is a set, then there exists
another set (the union of $x$) containing all the elements in each
element of $x$. The relativized version of Union asks to give a name
$\pi_a$ for each $a\in M[G]$ and proving $\val(G,\pi_a)=\union a$.
Let $\tau$ be the name for $a$, i.e.\ $a=\val(G,\tau)$; 
\citet{kunen2011set} gives $\pi_a$ in terms of $\tau$:
\begin{align*}
  \pi_a = \{\langle\theta,p \rangle &\in \dom(\union(\dom(\tau))) \times \PP : \\
&\exists \langle\sigma,q\rangle  \in \tau .
 \exists r \in \PP . \langle \theta,r\rangle \in \sigma \wedge
    p\leqslant r \wedge p \leqslant q \}
\end{align*}
Our formal definition is slightly different in order to ease the proof of
$\pi_a \in M$.  Since $\pi_a$ it is defined using Separation, one
needs to internalize $\pi_a$ as a formula
\isa{union{\isacharunderscore}name{\isacharunderscore}fm}. The
equation $\val(G,\pi_a)=\union a$ is proved by showing the mutual
inclusion; in both cases one uses that $G$ is a filter.
\begin{isabelle}
  \isacommand{lemma}\isamarkupfalse%
\ Union{\isacharunderscore}MG{\isacharunderscore}Eq\ {\isacharcolon}\ \isanewline
\ \ \isakeyword{assumes}\ {\isachardoublequoteopen}a\ {\isasymin}\ M{\isacharbrackleft}G{\isacharbrackright}{\isachardoublequoteclose}\ \isakeyword{and}\ {\isachardoublequoteopen}a\ {\isacharequal}\ val{\isacharparenleft}G{\isacharcomma}{\isasymtau}{\isacharparenright}{\isachardoublequoteclose}\ \isakeyword{and}\isanewline
\ \ \ \ \ \ \ \ \ \ {\isachardoublequoteopen}filter{\isacharparenleft}G{\isacharparenright}{\isachardoublequoteclose}\ \isakeyword{and}\ {\isachardoublequoteopen}{\isasymtau}\ {\isasymin}\ M{\isachardoublequoteclose}\isanewline
\ \ \isakeyword{shows}\ {\isachardoublequoteopen}{\isasymUnion}\ a\ {\isacharequal}\ val{\isacharparenleft}G{\isacharcomma}Union{\isacharunderscore}name{\isacharparenleft}{\isasymtau}{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}
\end{isabelle}
Since Union is absolute for any transitive class\footnote{This
  generalizes a result proved by Paulson.}
\begin{isabelle}
\isacommand{lemma}\isamarkupfalse%
\ Union{\isacharunderscore}abs{\isacharunderscore}trans\ {\isacharcolon}\ \isanewline
\ \ \isakeyword{assumes}\ {\isachardoublequoteopen}Transset{\isacharparenleft}Q{\isacharparenright}{\isachardoublequoteclose}\ \isakeyword{and}\ {\isachardoublequoteopen}a\ {\isasymin}\ Q{\isachardoublequoteclose}\ \isakeyword{and}\isanewline
\ \ \ \ \ \ \ \ \ \ {\isachardoublequoteopen}z\ {\isasymin}\ Q{\isachardoublequoteclose}\ \isakeyword{and}\ {\isachardoublequoteopen}{\isasymUnion}\ a\ {\isacharequal}\ z{\isachardoublequoteclose}\isanewline
\ \ \isakeyword{shows}\ {\isachardoublequoteopen}big{\isacharunderscore}union{\isacharparenleft}{\isacharhash}{\isacharhash}Q{\isacharcomma}a{\isacharcomma}z{\isacharparenright}{\isachardoublequoteclose}
\end{isabelle}
\noindent we may conclude that $M[G]$ is closed under Union:
\begin{isabelle}
  \isacommand{lemma}\isamarkupfalse%
  \ union{\isacharunderscore}in{\isacharunderscore}MG\ {\isacharcolon}\ \isanewline
  \ \ \isakeyword{assumes}\ {\isachardoublequoteopen}filter{\isacharparenleft}G{\isacharparenright}{\isachardoublequoteclose}\isanewline
\ \ \isakeyword{shows}\ {\isachardoublequoteopen}Union{\isacharunderscore}ax{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharbrackleft}G{\isacharbrackright}{\isacharparenright}{\isachardoublequoteclose}
\end{isabelle}

>>>>

The Union Axiom asserts that if $x$ is a set, then there exists
another set (the union of $x$) containing all the elements in each
element of $x$. The relativized version of Union asks to give a name
$\pi_a$ for each $a\in M[G]$ and proving $\val(G,\pi_a)=\union a$.
Let $\tau$ be the name for $a$, i.e.\ $a=\val(G,\tau)$; 
\citet{kunen2011set} gives $\pi_a$ in terms of $\tau$:
\begin{align*}
  \pi_a = \{\langle\theta,p \rangle &\in \dom(\union(\dom(\tau))) \times \PP : \\
&\exists \langle\sigma,q\rangle  \in \tau .
 \exists r \in \PP . \langle \theta,r\rangle \in \sigma \wedge
    p\leqslant r \wedge p \leqslant q \}
\end{align*}
Our formal definition is slightly different in order to ease the proof of
$\pi_a \in M$.  Since $\pi_a$ it is defined using Separation, one
needs to internalize $\pi_a$ as a formula
\isa{union{\isacharunderscore}name{\isacharunderscore}fm}. The
equation $\val(G,\pi_a)=\union a$ is proved by showing the mutual
inclusion; in both cases one uses that $G$ is a filter.
\begin{isabelle}
  \isacommand{lemma}\isamarkupfalse%
\ Union{\isacharunderscore}MG{\isacharunderscore}Eq\ {\isacharcolon}\ \isanewline
\ \ \isakeyword{assumes}\ {\isachardoublequoteopen}a\ {\isasymin}\ M{\isacharbrackleft}G{\isacharbrackright}{\isachardoublequoteclose}\ \isakeyword{and}\ {\isachardoublequoteopen}a\ {\isacharequal}\ val{\isacharparenleft}G{\isacharcomma}{\isasymtau}{\isacharparenright}{\isachardoublequoteclose}\ \isakeyword{and}\isanewline
\ \ \ \ \ \ \ \ \ \ {\isachardoublequoteopen}filter{\isacharparenleft}G{\isacharparenright}{\isachardoublequoteclose}\ \isakeyword{and}\ {\isachardoublequoteopen}{\isasymtau}\ {\isasymin}\ M{\isachardoublequoteclose}\isanewline
\ \ \isakeyword{shows}\ {\isachardoublequoteopen}{\isasymUnion}\ a\ {\isacharequal}\ val{\isacharparenleft}G{\isacharcomma}Union{\isacharunderscore}name{\isacharparenleft}{\isasymtau}{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}
\end{isabelle}

Since Union is absolute for any transitive class
\medskip

\fbox{No generaliza a Paulson porque \'el prueba el sii.}

\fbox{Y capaz su prueba de dos l\'{\i}neas tambi\'en lo hace}

\fbox{ac\'a (como Fundaci\'on), as\'{\i} que cuidado.}
\begin{isabelle}
\isacommand{lemma}\isamarkupfalse%
\ Union{\isacharunderscore}abs{\isacharunderscore}trans\ {\isacharcolon}\ \isanewline
\ \ \isakeyword{assumes}\ {\isachardoublequoteopen}Transset{\isacharparenleft}Q{\isacharparenright}{\isachardoublequoteclose}\ \isakeyword{and}\ {\isachardoublequoteopen}a\ {\isasymin}\ Q{\isachardoublequoteclose}\ \isakeyword{and}\isanewline
\ \ \ \ \ \ \ \ \ \ {\isachardoublequoteopen}z\ {\isasymin}\ Q{\isachardoublequoteclose}\ \isakeyword{and}\ {\isachardoublequoteopen}{\isasymUnion}\ a\ {\isacharequal}\ z{\isachardoublequoteclose}\isanewline
\ \ \isakeyword{shows}\ {\isachardoublequoteopen}big{\isacharunderscore}union{\isacharparenleft}{\isacharhash}{\isacharhash}Q{\isacharcomma}a{\isacharcomma}z{\isacharparenright}{\isachardoublequoteclose}
\end{isabelle}
\noindent we may conclude that $M[G]$ is closed under Union:
\begin{isabelle}
  \isacommand{lemma}\isamarkupfalse%
  \ union{\isacharunderscore}in{\isacharunderscore}MG\ {\isacharcolon}\ \isanewline
  \ \ \isakeyword{assumes}\ {\isachardoublequoteopen}filter{\isacharparenleft}G{\isacharparenright}{\isachardoublequoteclose}\isanewline
\ \ \isakeyword{shows}\ {\isachardoublequoteopen}Union{\isacharunderscore}ax{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharbrackleft}G{\isacharbrackright}{\isacharparenright}{\isachardoublequoteclose}
\end{isabelle}

### 2018-12-19 LICS/hacking.tex
ed. varias, esp. Hacking y comentario para Miguel

There are several lemmas that were declared, in later developments, as
introduction/simplification rules (notably, the rule
\isatt{equalityI}). They raised a warning, and we have 
eliminated some of them, but in some cases we had to keep them because
the proof works because it is insisted that the rule is \emph{safe}.

The most notable changes, located in the theory \texttt{Relative}, are
the following:
\begin{enumerate}
\item\label{item:1} We eliminated the requirement that the relative Axiom of Replacement
  is satisfied by a class $M$ to be in the locale \isatt{M\_trivial}. 
\item\label{item:2} We moved the requirement of the Powerset Axiom to \isatt{M\_basic}. 
\item\label{item:3} We replaced the need that the set of natural numbers is in $M$ by the
  milder hypothesis that $M(0)$.
\end{enumerate}

@@

The proof, for instance, that the constructible universe $L$ satisfies
the modified locale \isatt{M\_trivial} holds with minor
modifications. Nevertheless, in order to have a neater presentation,
we have stripped off several sections concerning $L$ from the theories
\isatt{L\_axioms} and \isatt{Internalize}, and we merged them to form
the new file  \isatt{Internalizations}. Actually we believe that the
\isatt{ZF-Constructible} session  would benefit from some changes in
the way some results are organized accross theory files; for
intance, by cataloging all or most of the internalized formulas in one
file.

>>>>

In \cite{paulson_2003}, Paulson presented his formalization of the
relative consistency of the Axiom of Choice. This development is
included inside the Isabelle distribution with session name
\isatt{ZF-Constructible}. The main technical devices, invented by
G\"odel for this purpose, are \emph{relativization} and
\emph{absoluteness}. In a nutshell, to relativize a formula $\phi$ to
a class $C$, it is enough to restrict its quantifiers to $C$. The
example of \isatt{upair\_ax} in
Section~\ref{sec:axioms-models-set-theory}, the relativized version of
the Pairing Axiom, is extracted from \texttt{Relative}, one of the
core theories of \isatt{ZF-Constructible}. On the other hand, $\phi$
is \emph{absolute} for $C$ if it is equivalent to its relativization,
meaning that the statement made $\phi$ coincides with what $C$
``believes'' $\phi$ is saying. Paulson shows that under certain
hypothesis  on a class $M$ (condensed in the locale \isatt{M\_trivial}), a plethora of
absoluteness and closure results can be proved about $M$.

The development of forcing, and the study of ctms in general, takes
absoluteness as a starting point. We were not able to work with
\isatt{ZF-Constructible} right out-of-the-box. The main reason is that
we can't expect state the ``class version'' of Replacement for a
\emph{set} $M$ by
using first-order formulas, since predicates \isatt{P::"i=>o"} can't
be proved to be only the definable ones. Therefore, we had to make
some modifications to the various sets of hypothesis in several
locales to make the results available as tools for the present and
future developments.

%% There are several lemmas that were declared, in later developments, as
%% introduction/simplification rules (notably, the rule
%% \isatt{equalityI}). They raised a warning, and we have 
%% eliminated some of them, but in some cases we had to keep them because
%% the proof works because it is insisted that the rule is \emph{safe}.

The most notable changes, located in the theory \texttt{Relative}, are
the following:
\begin{enumerate}
\item\label{item:1} We eliminated the requirement that the relative Axiom of Replacement
  is satisfied by a class $M$ to be in the locale \isatt{M\_trivial}. 
\item\label{item:2} We moved the requirement of the Powerset Axiom to \isatt{M\_basic}. 
\item\label{item:3} We replaced the need that the set of natural numbers is in $M$ by the
  milder hypothesis that $M(0)$. Actually, most results should follow
  by only assuming that $M$ is nonempty.
\end{enumerate}

@@

The proof, for instance, that the constructible universe $L$ satisfies
the modified locale \isatt{M\_trivial} holds with minor
modifications. Nevertheless, in order to have a neater presentation,
we have stripped off several sections concerning $L$ from the theories
\isatt{L\_axioms} and \isatt{Internalize}, and we merged them to form
the new file  \isatt{Internalizations}.

\medskip
\fbox{>>>Poner lo que sigue en las conclusiones???}
\medskip

We believe that the
\isatt{ZF-Constructible} session  would benefit from some changes in
the way some results are organized accross theory files; for
intance, by cataloging all or most of the internalized formulas in one
file. Another, most basic example would be to start out with an even
more locale that only assumes $M$ to be a nonempty transitive class,
as many absoluteness results follow from these hypothesis.  
It is our desire to advocate a future work to a thorough
revision of the development of constructibility to maximize modularity.
